# Basic tools {#basics}

In this Chapter, we present basic tools that will be important when interacting with big data systems, most importantly the command-line interface (CLI) in a Unix shell and several utilities (`less`, `vi`, `awk` and `make`).

## Command line tools

We assume some familiarity with the Unix shell, for example as in http://swcarpentry.github.io/shell-novice/.

We also assume that you have access to a shell, either because you use Linux or OS X or because you have the right tools on Windows (for example [Cygwin](https://www.cygwin.com/) or the Bash shell in Windows 10).

### Why use the command line?

- Batch processing

- Cluster and cloud computing

### Basic Unix commands

### Useful tools

#### `less`

#### `vi`

#### `awk`

### Example

Let's apply some of the techniques described in @Blackwell2012.

## Makefiles

## Git and GitHub

Guest lecture by [Ista Zahn](https://www.iq.harvard.edu/people/ista-zahn).

## Buffer

```{r}
ratings <- read.csv("~/Dropbox/data17/ml-latest-small/ratings.csv")
head(ratings)
```
```{r}
links <- read.csv("~/Dropbox/data17/ml-latest-small/links.csv")
head(links)
```

```{r}
movies <- read.csv("~/Dropbox/data17/ml-latest-small/movies.csv")
head(movies)
```
```{r}
tags <- read.csv("~/Dropbox/data17/ml-latest-small/tags.csv")
head(tags)
```

