# Optimization {#optimization}

In this Chapter, we will see how to measure and improve code performance.

## Measuring performance

### Benchmarking

Reading: http://adv-r.had.co.nz/Performance.html#microbenchmarking

There are several ways to benchmark code (see http://www.alexejgossmann.com/benchmarking_r/) from `system.time` to dedicated packages such as `rbenchmark` (@rbenchmark) or `microbenchmark` (@microbenchmark).

Let's start with an example from @Wickham2014.

```{r}
library(microbenchmark)
m <- microbenchmark(
  times = 1000, # default is 100
  "[32, 11]"      = mtcars[32, 11],
  "$carb[32]"     = mtcars$carb[32],
  "[[c(11, 32)]]" = mtcars[[c(11, 32)]],
  "[[11]][32]"    = mtcars[[11]][32],
  ".subset2"      = .subset2(mtcars, 11)[32]
)
m
```


```{r}
ggplot2::autoplot(m)
```


### Profiling and optimization

Reading: http://adv-r.had.co.nz/Profiling.html#measure-perf

Let's compare three ways of estimating a linear regression: with built-in `lm` and with two functions we defined in package `Linreg` in Chapter \@ref(packages).

```{r echo=FALSE}
suppressPackageStartupMessages(library(Linreg))
```

```{r}
library(Linreg)
data(cats, package = "MASS")
fit1 <- lm(Hwt ~ Bwt, data = cats)
fit2 <- linmod(Hwt ~ Bwt, data = cats)
fit3 <- linmodEst(cbind(1, cats$Bwt), cats$Hwt)
all.equal(round(coef(fit1), 5), round(coef(fit2), 5))
all.equal(round(coef(fit1), 5), round(fit3$coefficients, 5), check.names = FALSE)

m <- microbenchmark(
  fit1 <- lm(Hwt ~ Bwt, data = cats),
  fit2 <- linmod(Hwt ~ Bwt, data = cats),
  fit3 <- linmodEst(cbind(1, cats$Bwt), cats$Hwt)
  # custom checks can be performed with the 'check' argument
)
m
ggplot2::autoplot(m)
```


## Improving performance

- Vectorize

- Parallelize

- Use a faster language (C/C++, Fortran, ...)

- Use different tools (as in Chapter \@ref(bigdata))

## Vectorization

Let's take an example from a blog post (that seems to be [gone](http://www.babelgraph.org/wp/?p=358)).  It's used in @Wickham2014 [, Section [Case studies](http://adv-r.had.co.nz/Rcpp.html#rcpp-case-studies)].

```{r}
vacc1a <- function(age, female, ily) {
  p <- 0.25 + 0.3 * 1 / (1 - exp(0.04 * age)) + 0.1 * ily
  p <- p * if (female) 1.25 else 0.75
  p <- max(0, p)
  p <- min(1, p)
  p
}

set.seed(1959)
n <- 1000
age <- rnorm(n, mean = 50, sd = 10)
female <- sample(c(T, F), n, rep = TRUE)
ily <- sample(c(T, F), n, prob = c(0.8, 0.2), rep = TRUE)

vacc1a(age[1], female[1], ily[1])
vacc1a(age[2], female[2], ily[2])
vacc1a(age[3], female[3], ily[3])
```

`vacc1a` is not designed for vector inputs

```{r}
vacc1a(age, female, ily)
```

It should be called

```{r}
vacc1a(age[1], female[1], ily[1])
vacc1a(age[2], female[2], ily[2])
vacc1a(age[3], female[3], ily[3])

```

We can use a loop:

```{r}
out <- numeric(n)
for (i in 1:n)
  out[i] <- vacc1a(age[i], female[i], ily[i])
```

or one of the `apply` functions:

```{r}
vacc0<- function(age, female, ily) {
  sapply(1:n, function(i) vacc1a(age[i], female[i], ily[i]))
}

out0 <- vacc0(age, female, ily)
```

```{r}
all.equal(out, out0)
```

But, it's convenient for the function to support vector inputs, instead of relying on users writing their own wrappers.  We can loop inside the function body.

```{r}
vacc1 <- function(age, female, ily) {
  n <- length(age)
  out <- numeric(n)
  for (i in seq_len(n)) {
    out[i] <- vacc1a(age[i], female[i], ily[i])
  }
  out
}
```

or we can rely on base R functions that accept vector inputs

```{r}
vacc2 <- function(age, female, ily) {
  p <- 0.25 + 0.3 * 1 / (1 - exp(0.04 * age)) + 0.1 * ily
  p <- p * ifelse(female, 1.25, 0.75)
  p <- pmax(0, p)
  p <- pmin(1, p)
  p
}
```

## Parallelization

```{r}
library(parallel)
cores <- detectCores()
cores
```

```{r}
vacc3 <- function(age, female, ily) {
  mcmapply(function(i) vacc1a(age[i], female[i], ily[i]), 1:n, mc.cores = cores - 1)
}

out3 <- vacc3(age, female, ily)
```

```{r}
library(microbenchmark)
m <- microbenchmark(
  vacc0 = vacc0(age, female, ily),
  vacc1 = vacc1(age, female, ily),
  vacc2 = vacc2(age, female, ily),
  vacc3 = vacc3(age, female, ily)
)
m
ggplot2::autoplot(m)
```

So, what's going on?

We will talk more about parallelization tools and techniques in Chapter `@ref(bigdata).

## Introduction to C++

- C++ is a very powerful object-oriented language.

- Many tutorials are available on-line, for example http://www.cplusplus.com/doc/tutorial/.

- R is _intepreted_, C++ is _compiled_ and typically much faster (in loops for examples).

- Our introduction to C++ is from an R perspective.  Python (and most interpreted languages) can be extended with C++ too.

### Rcpp

Reading: http://adv-r.had.co.nz/Rcpp.html

- `Rcpp` @Eddelbuettel2013 makes it very easy to use C++ code in R (for example to speed up a function or to wrap methods already implemented in C++).

- `Rcpp` provides "syntactic sugar" that makes is easy to leverage C++ even without a deep knowledge of it.

- To use `Rcpp`, you need a C++ compiler:
    - Windows: [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
    - OS X: [Xcode](https://developer.apple.com/xcode/)
    - Linux: `r-base-dev` from package manager
    
### Hello World!

```{r}
library(Rcpp)
cppFunction('void hello(){
  Rprintf("Hello, world!");
}')
hello
hello()
```

`Rprintf` is the counterpart of C++ [`printf`](http://www.cplusplus.com/reference/cstdio/printf/) function.

Let's take the first example of @Wickham2014, Section [Getting started with C++](http://adv-r.had.co.nz/Rcpp.html#rcpp-intro).

```{r eval=FALSE}
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')
```

We have to specify the input type and the output type.  As expected

```{r eval=FALSE}
add(1, 2, 3)
```

returns 6. How about?

```{r eval=FALSE}
add(1.1, 2.2, 3.3)
```

```{r eval=FALSE}
cppFunction('double addd(double x, double y, double z) {
  double sum = x + y + z;
  return sum;
}')
```

With `addd` we do get 6.6:

```{r eval=FALSE}
addd(1.1, 2.2, 3.3)
```

### `sourceCpp`

When C++ code takes more than a couple of lines, it's more convenient to create a stand-alone C++ source file.

From the RStudio default template:

```{r engine="cpp", eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

NumericVector timesTwo(NumericVector x) {
  return x * 2;
}

/*** R
timesTwo(42)
*/
```

From R, we can use `sourceCpp` to access `timesTwo` in R:

```{r eval=FALSE}
sourceCpp("src/times-two.cpp")
timesTwo(100)
```

### Data types

`int`
`double`
`bool`
`string`

`NumericVector`
`LogicalVector`
`IntegerVector`
`CharacterVector`

`NumericMatrix`
`IntegerMatrix`
`LogicalMatrix`
`CharacterMatrix`

`NA_REAL`
`NA_INTEGER`
`NA_STRING`
`NA_LOGICAL`

`List`
`DataFrame`
`Function`

...

### Sugar

Reading: http://adv-r.had.co.nz/Rcpp.html#rcpp-sugar.

- Vectorization of `+`, `*`, `-`, `/`, `pow`, `<`, `<=`, `>`, `>=`, `==`, `!=`, `!`

-x`


```{exercise}
Can you write an `Rcpp` function similar to `addd` but accepting vector arguments?
```

```{r}
cppFunction('NumericVector addv(NumericVector x, NumericVector y, NumericVector z) {
  NumericVector sum = x + y + z;
  return sum;
}')
```

### Example (continued)

```{r engine="cpp", eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

double vacc3a(double age, bool female, bool ily){
  double p = 0.25 + 0.3 * 1 / (1 - exp(0.04 * age)) + 0.1 * ily;
  p = p * (female ? 1.25 : 0.75);
  p = std::max(p, 0.0);
  p = std::min(p, 1.0);
  return p;
}

// [[Rcpp::export]]
NumericVector vacc3(NumericVector age, LogicalVector female, 
                    LogicalVector ily) {
  int n = age.size();
  NumericVector out(n);

  for(int i = 0; i < n; ++i) {
    out[i] = vacc3a(age[i], female[i], ily[i]);
  }

  return out;
}
```

### Back to Linreg

- `armadillo` is a very powerful C++ linear algebra library: http://arma.sourceforge.net/

- It can be used in `Rcpp` via the `RcppArmadillo` package.

```{exercise}
Can you write an `Rcpp` function similar to `linmodEst`?
```

```{r eval=TRUE, eval=FALSE}
linmodEst <- function(x, y) {
  ## CC: crossprod or a QR decomposition (as in the original version) are more efficient
  coef <- solve(t(x) %*% x) %*% t(x) %*% y
  ## degrees of freedom and standard deviation of residuals
  df <- nrow(x) - ncol(x)
  sigma2 <- sum((y - x %*% coef) ^ 2) / df
  ## compute sigma^2 * (x’x)^-1
  vcov <- sigma2 * solve(t(x) %*% x)
  colnames(vcov) <- rownames(vcov) <- colnames(x)
  list(
    coefficients = coef,
    vcov = vcov,
    sigma = sqrt(sigma2),
    df = df
  )
}
```

## Rcpp packages

Readings:
- https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-package.pdf
- http://adv-r.had.co.nz/Rcpp.html#rcpp-package
